\chapter{Solution Description}
\label{chap:description}

This chapter will present the design and implementation of a solution
for the problem stated in Chapter \ref{chap:motivation}.
The solution has been designed according to the previously stated requirements.


\section{Overview}
From an architectural point of view
the solution consists of the following components:
\begin{description}
\item[Name mangling]
	This component is responsible for generating the \ac{TTCN-3} types
	(e.g.\ signatures, object types, port types)
	used in writing testcases.
	At runtime it will perform a reverse mangling
	between the types used in the abstract test suite
	and the original Java types.
\item[Codec]
	This component implements the standard functionality of
	encoding (and decoding) between
	\ac{TTCN-3} types and the corresponding Java types.
	In addition the Codec is responsible for storing the Java objects
	(since they are stateful and must survive
	between method invocations).
	The Codec encodes/decodes all parameters for signature calls
	(the parameters may be primitive values or object references)
	as well as thrown exceptions.
\item[Port Plugin]
	A \ac{TTCN-3} test system consists of several parallel test components.
	Each component may have several ports, and each port can be used
	to send or receive certain types of messages
	(or to make or receive calls for certain signatures).
	The abstract communication model
	requires ports to be connected to one another before being used.
	In this project,
	whenever a (generated) signature is called through a port,
	the Port Plugin is asked to handle this call.
	It receives the respective signature and the encoded parameters
	and is responsible for invoking the actual Java method requested.
	It also enqueues returned values or thrown exceptions
	for processing by the testcase code.
\end{description}

The implementation language is Java.
By using the \verb=reflection= package
an examination of compiled code is made possible.
This permits the automatic generation of corresponding \ac{TTCN-3} types
for any compiled Java types (e.g.\ classes, interfaces, enumerations).
The codec and port plugin implement standard test system interfaces.
The whole project is compiled as a plugin for
Testing Technologies' TTworkbench \citep{website:ttworkbench}
(a test development and execution environment
based on the Eclipse \citep{website:eclipse} \ac{IDE}).

\input{./tex/description-type-mapping.tex}

\input{./tex/description-name-mangling.tex}

\input{./tex/description-codec.tex}

\input{./tex/description-port-plugin.tex}

\begin{comment}
\begin{verbatimtab}
General overview
	TODO: add workflow diagram
Type mapping
	Primitive types
	Using Java Objects from TTCN-3 code
	Mapping Java Object Types
		What output should look like

Name mangling
	Escaping special characters
	Avoiding name-clashes for overloaded methods
	Avoiding name-clashes between generated and user-defined identifiers
Storage Codec
	Registry architecture
Port Plugin
Limitations
\end{verbatimtab}
\end{comment}
